Penalized regression for left-truncated and right-censored survival data

### 2021-04-16 manuscript.tex
unbold text, add last name
de-pluralize Data Sciences (author affils)
Clarifications to the data generating process

%\author[1]{Sarah McGough\footnote{Contributed equally.}}
\newcommand*\samethanks[1][\value{footnote}]{\footnotemark[#1]}

\author[1]{Sarah F. McGough\thanks{Contributed equally.}}
\author[1]{Devin Incerti\samethanks}
\author[1]{Svetlana Lyalina}
\author[1]{Ryan Copping}
\author[2, 3]{Balasubramanian Narasimhan}
\author[2, 3]{Robert Tibshirani}
\affil[1]{Genentech, Inc, South San Francisco, CA, USA}
\affil[2]{Department of Statistics, Stanford University, Stanford, CA, USA}
\affil[3]{Department of Biomedical Data Sciences, Stanford University, Stanford, CA, USA}

@@

\section{Simulation study} \label{sec:simulation}
Simulations were conducted to assess the performance of the method with LTRC data. To set notation, let $T_i$, $U_i$, and $V_i$ be latent survival, right censoring, and study entry times for patient $i$, respectively. The simulation proceeds by simulating $T_i$, $U_i$, and $V_i$ from suitable survival distributions and assumes that $T_i$, $U_i$, and $V_i$ are uncorrelated. The observed survival time is given by $Y_i = \min(T_i, U_i)$ and a patient is right-censored if $T_i > U_i$. Patients are left-truncated (i.e., hidden from the sample) if $V_i > Y_i$.

@@

with the $k$th binary variable $\tilde{x}_{i,k}=1$ if $\tilde{x}_{i,k}^{*} > 0$ and $0$ otherwise. Each diagonal element of $\Sigma$ equals 1 and non-diagonal elements determine the correlation between variables. The probability, $\pi_k$ that  variable $k$ was equal to $1$ was randomly sampled from a uniform distribution ranging from .2 to .8; that is, we sampled $\pi_k \sim U(0.2, 0.8)$ and set $\mu_k = \Phi^{-1}(\pi_k)$ where $\Phi(\cdot)$ is the cumulative density function of the normal distribution. $\Sigma$ was generated by filling a $\tilde{p} \times \tilde{p}$ matrix, $\tilde{\Sigma}$, with $N(0,1)$ random variables and setting  $\Sigma = S/D^TD$ with $S = \tilde{\Sigma}^T\tilde{\Sigma}$ and $D$ a vector containing the diagonal elements of $S$.

Latent survival time was simulated from a proportional hazards Weibull distribution with shape parameter $a$ and scale parameter $m_i$ based on evidence from the CGDB as detailed in \autoref{subsec:sim-calibration}. Predictors predict latent survival time through their effect on the scale parameter,

@@

where $\alpha$ is an intercept and $\beta$ is a vector of coefficients. $\alpha$ and values of $\beta$ for the CGDB variables were estimated from the CGDB data. The probability that the coefficient of a binary variables was non-zero was $0.5$ and drawn from a Bernoulli distribution. Non-zero coefficients were drawn from a uniform distribution with lower bound of $-0.25$ and upper bound of $0.25$ so that the minimum and maximum hazard ratios were $0.78$ and $1.28$, respectively.

@@

The CGDB variables were standardized to have mean $0$ and standard deviation $1$ and coefficients were estimated using a Cox proportional hazards model. The full coefficient vector combined the coefficients from the CGDB variables with the coefficients from the simulated binary variables. The scale parameter was set to $m = \alpha + x^T\beta - E(x^T\beta)$ so that predicted survival averaged across patients was equal to predicted survival from the intercept only Weibull model.

@@

\section*{Acknowledgments} 
We thank T. Hastie and K. Tay for their work in implementing left truncation-adjusted Cox models in \pkg{glmnet}. 
\textbf{Competing interests}: R.T. and B.N. are paid consultants for Roche.

>>>>

\newcommand*\samethanks[1][\value{footnote}]{\footnotemark[#1]}

\author[1]{Sarah F. McGough\thanks{Contributed equally.}}
\author[1]{Devin Incerti\samethanks}
\author[1]{Svetlana Lyalina}
\author[1]{Ryan Copping}
\author[2, 3]{Balasubramanian Narasimhan}
\author[2, 3]{Robert Tibshirani}
\affil[1]{Genentech, Inc, South San Francisco, CA, USA}
\affil[2]{Department of Statistics, Stanford University, Stanford, CA, USA}
\affil[3]{Department of Biomedical Data Science, Stanford University, Stanford, CA, USA}

@@

\section{Simulation study} \label{sec:simulation}
Simulations were conducted to assess the performance of the method with LTRC data. To set notation, let $T_i$, $U_i$, and $V_i$ be latent survival, right censoring, and study entry times for patient $i$, respectively. The simulation proceeds by simulating $T_i$, $U_i$, and $V_i$ from suitable survival distributions and assumes that $T_i$, $U_i$, and $V_i$ are uncorrelated. The observed survival time is given by $Y_i = \min(T_i, U_i)$ and a patient is right-censored if $T_i > U_i$. Patients are left-truncated (i.e., excluded from the sample) if $V_i > Y_i$.

@@

with the $k$th binary variable $\tilde{x}_{i,k}=1$ if $\tilde{x}_{i,k}^{*} > 0$ and $0$ otherwise. Each diagonal element of $\Sigma$ equals 1 and non-diagonal elements determine the correlation between variables. The probability, $\pi_k$ that  variable $k$ was equal to $1$ was randomly sampled from a uniform distribution ranging from .2 to .8; that is, we sampled $\pi_k \sim U(0.2, 0.8)$ and set $\mu_k = \Phi^{-1}(\pi_k)$ where $\Phi(\cdot)$ is the cumulative density function of the normal distribution. $\Sigma$ was generated by first constructing a positive semi-definite matrix $S = \tilde{\Sigma}^T\tilde{\Sigma}$ by filling the $\tilde{p} \times \tilde{p}$ matrix, $\tilde{\Sigma}$, with $N(0,1)$ random variables. $S$ was then scaled into a correlation matrix by setting $\Sigma = S/D^TD$ where $D$ is a vector containing the diagonal elements of $S$.

Latent survival time was simulated from a  Weibull distribution based on evidence from the CGDB as detailed in \autoref{subsec:sim-calibration}. We used a proportional hazards parameterization of the Weibull distribution with shape parameter $a$, scale parameter $m$, and for $t> 0$, probability density function given by,

\begin{equation}
f(t) = amt^{a-1}e^{-mt^a}.
\end{equation}

The predictors were used to model latent survival time through their effect on the scale parameter,

@@

where $\alpha$ is an intercept and $\beta$ is a vector of coefficients. $\alpha$ and values of $\beta$ for the CGDB variables were estimated from the CGDB data as described in \autoref{subsec:sim-calibration}. The probability that the coefficient of a binary variables was non-zero was $0.5$ and drawn from a Bernoulli distribution. Non-zero coefficients were drawn from a uniform distribution with lower bound of $-0.25$ and upper bound of $0.25$ so that the minimum and maximum hazard ratios were $0.78$ and $1.28$, respectively.

@@

The CGDB variables were standardized to have mean $0$ and standard deviation $1$ and coefficients were estimated using a Cox proportional hazards model. The full coefficient vector combined the coefficients from the CGDB variables with the coefficients from the simulated binary variables. The natural logarithm of the scale parameter was set to $\log(m_i) = \alpha + x^T_i\beta - E(x^T_i\beta)$ so that predicted survival averaged across patients was equal to predicted survival from the intercept only Weibull model.

@@

\section*{Acknowledgments} 
We thank T. Hastie and K. Tay for their work in implementing left truncation-adjusted Cox models in \pkg{glmnet}. 
Competing interests: R. Tibshirani and B. Narasimhan are paid consultants for Roche.
